---
title: "ITS WORKFLOW-QIIME2"
author: "Steph"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    theme: readable
    highlight: tango
    toc: true
    toc_depth: 3
    toc_float: true
---
# I. ITS WORKFLOW-QIIME2

#### **Script to process ITS sequences from LMNP and IPNP**


## 1. IMPORTING AND DEMULTIPLEXING SEQUENCES


### IMPORTING TO QIIME2

- Make directories for each library

```
mkdir ITS1
mkdir ITS2
mkdir ITS3
mkdir ITS4
mkdir ITS5
mkdir ITS6
mkdir ITS7
mkdir ITS8
mkdir ITS9

```

- Loop to rename forward seqs

```
for i in ITS* ;do mv ITS*/ITS_*1.fastq.gz  ITS*/forward.fastq.gz ; done
```

- Loop to rename reverse seqs

```
for i in ITS* ;do mv ITS*/ITS_*2.fastq.gz  ITS*/reverse.fastq.gz ; done
```

- Loop to import files to QIIME2

```
for i in ITS*; do qiime tools import --type MultiplexedPairedEndBarcodeInSequence --input-path $i --output-path $i.qza ; done

```
### DEMULTIPLEXING WITH CUTADAPT

```
qiime cutadapt demux-paired\
  --i-seqs its1.qza\
  --m-forward-barcodes-file ../../maps_its/its1.txt\
  --m-forward-barcodes-column BarcodeSequence\
  --output-dir its1_demux

qiime cutadapt demux-paired\
  --i-seqs its2.qza\
  --m-forward-barcodes-file ../../maps_its/its2.txt\
  --m-forward-barcodes-column BarcodeSequence\
  --output-dir its2_demux

qiime cutadapt demux-paired\
  --i-seqs its3.qza\
  --m-forward-barcodes-file ../../maps_its/its3.txt\
  --m-forward-barcodes-column BarcodeSequence\
  --output-dir its3_demux

qiime cutadapt demux-paired\
  --i-seqs its4.qza --m-forward-barcodes-file ../../maps_its/its4.txt\
  --m-forward-barcodes-column BarcodeSequence\
  --output-dir its4_demux

qiime cutadapt demux-paired  --i-seqs its5.qza\
  --m-forward-barcodes-file ../../maps_its/its5.txt\
  --m-forward-barcodes-column BarcodeSequence\
  --output-dir its5_demux

qiime cutadapt demux-paired  --i-seqs its6.qza\
  --m-forward-barcodes-file ../../maps_its/its6.txt\
  --m-forward-barcodes-column BarcodeSequence\
  --output-dir its6_demux

qiime cutadapt demux-paired  --i-seqs its7.qza\
  --m-forward-barcodes-file ../../maps_its/its7.txt\
  --m-forward-barcodes-column BarcodeSequence\
  --output-dir its7_demux

qiime cutadapt demux-paired\
  --i-seqs its8.qza\
  --m-forward-barcodes-file ../../maps_its/its8.txt\
  --m-forward-barcodes-column BarcodeSequence\
  --output-dir its8_demux

qiime cutadapt demux-paired\
  --i-seqs its9.qza\
  --m-forward-barcodes-file ../../maps_its/its9.txt\
  --m-forward-barcodes-column BarcodeSequence\
  --output-dir its9_demux

```

## 2. ITSXPRESS

### OPTION 1: ITSXPRESS EN QIIME2 (TESTING)

- ITSxpress in QIIME2

```
for i in its*_demux ; do qiime itsxpress trim-pair --i-per-sample-sequences $i/per_sample_sequences.qza --p-region ITS2 --p-taxa F --o-trimmed itsxpress_2024_$i;done

for i in its*_demux ; do qiime itsxpress trim-pair-output-unmerged --i-per-sample-sequences $i/per_sample_sequences.qza --p-region ITS2 --p-taxa F --o-trimmed itsxpress_2024_unmerged$i;done


```

- Exporting and importing demultiplexed

```
for i in itsxpress_2024_its*; do qiime tools export --input-path $i --output-path exported; done
```

```
qiime tools import  --type 'SampleData[SequencesWithQuality]'  --input-path manifest.txt  --output-path single-end-demux-qiime2.qza  --input-format SingleEndFastqManifestPhred33V2
```


### OPTION 2: STANDALONE ITSXPRESS

- Exporting data

```
for i in its*_demux; do qiime tools export  --input-path  $i/per_sample_sequences.qza --output-path  exported_$i; done

```

- Move all data to one directory and run this bash script (*chatGPT*):

```{bash, eval=FALSE}

#!/bin/bash

# Set the path to the itsxpress executable
ITSXPRESS_EXECUTABLE="itsxpress"

# Loop through the files
for forward_file in *_R1_001.fastq.gz; do
    # Extract the sample name from the forward file
    sample_name=$(basename "$forward_file" _R1_001.fastq.gz)

    # Construct the reverse file name
    reverse_file="${sample_name}_R2_001.fastq.gz"

    # Run itsxpress command
    $ITSXPRESS_EXECUTABLE --fastq "$forward_file" --fastq2 "$reverse_file" --region ITS2 \
        --taxa Fungi --log "${sample_name}_logfile.txt" --outfile "${sample_name}_trimmed_reads.fastq.gz" --threads 2

    # Optionally, you can print a message indicating the completion of each iteration
    echo "Processing $forward_file and $reverse_file"
done
```

- Import to QIIME2

```
qiime tools import\
  --type 'SampleData[SequencesWithQuality]'\
  --input-path manifest.txt\
  --output-path single-end-demux-standalone.qza\
  --input-format SingleEndFastqManifestPhred33V2
```

## 2. ASV's APPROACH

- Run DADA2 on the two data-sets. 

```
 for i in *; do qiime dada2 denoise-single --i-demultiplexed-seqs $i --p-trunc-len 0 --p-n-threads 4 --output-dir dada2_$i; done
```

- Filter seqs by length (>50bp)

```{bash, eval=FALSE}
qiime feature-table filter-seqs\
  --m-metadata-file dada2_single-end-demux-qiime2.qza/representative_sequences.qza\
  --i-data dada2_single-end-demux-qiime2.qza/representative_sequences.qza\
  --o-filtered-data representative_sequences_50_qiime2.qza\
  --p-where "length(sequence) > 49"

qiime feature-table filter-seqs\
  --m-metadata-file dada2_single-end-demux-standalone.qza/representative_sequences.qza\
  --i-data dada2_single-end-demux-standalone.qza/representative_sequences.qza\
  --o-filtered-data representative_sequences_50_standalone.qza\
  --p-where "length(sequence) > 49"

```

- Filter table based on seqs

```{bash, eval=FALSE}
qiime feature-table filter-features\
  --i-table dada2_single-end-demux-qiime2.qza/table.qza\
  --m-metadata-file representative_sequences_50_qiime2.qza\
  --o-filtered-table table_50_qiime2.qza

qiime feature-table filter-features\
  --i-table dada2_single-end-demux-standalone.qza/table.qza\
  --m-metadata-file representative_sequences_50_standalone.qza\
  --o-filtered-table table_50_standalone.qza
```





## 3. OTU's APPROACH

For qiime2 data, we have to export and run:

```
for i in *.fastq.gz; do reformat in=$i out=clean_$i minconsecutivebases=1; done
```

Then import again and run all again. This is due to an error of ITSxpress that generates sequences with 0 lentgh. The next steps will be run for the standalone and qiime2 data to custer in order to cluster to OTUS.


- Filter by q-score

```
for i in *;do  qiime quality-filter q-score --i-demux $i --output-dir filterqscore_$i; done

```

- Derreplication

```
for i in filterqscore_*; do qiime vsearch dereplicate-sequences --i-sequences $i/filtered_sequences.qza --output-dir derep_$i;done
```


- Clustering *de novo*

```
for i in derep_* ; do qiime vsearch cluster-features-de-novo --i-sequences $i/dereplicated_sequences.qza --i-table $i/dereplicated_table.qza --p-perc-identity 0.97 --p-threads 4 --output-dir cluster97_$i; done
```

- Chimera checking and filter from table

```
for i in cluster97_*; do qiime vsearch uchime-denovo --i-sequences $i/clustered_sequences.qza --i-table $i/clustered_table.qza --output-dir chimera97_$i;done

```

```
qiime feature-table filter-features\
  --i-table cluster97_derep_filterqscore_single-end-demux-qiime2-reformat.qza/clustered_table.qza\
  --m-metadata-file chimera97_cluster_derep_filterqscore_single-end-demux-qiime2-reformat.qza/nonchimeras.qza  \
  --o-filtered-table chimera97_cluster_derep_filterqscore_single-end-demux-qiime2-reformat.qza/table97-nonchimeras-qiime2.qza

```

```
qiime feature-table filter-features\
--i-table cluster97_derep_filterqscore_single-end-demux-standalone.qza/clustered_table.qza\
--m-metadata-file chimera97_cluster_derep_filterqscore_single-end-demux-standalone.qza/nonchimeras.qza  --o-filtered-table chimera97_cluster_derep_filterqscore_single-end-demux-standalone.qza/table97-nonchimeras-standalone.qza
```



**Filtering singletons from table and seqs** (voy por aquí)

```
 for i in chimera* ; do qiime feature-table filter-features --i-table $i/table*.qza --p-min-frequency 2 --o-filtered-table $i/filtered_$i ; done
 ```
 
``` 
for i in chimera*/; do
    data_file=$(find $i -name 'nonchimeras*' -type f)
    table_file=$(find $i -name 'filtered-table*' -type f)

    qiime feature-table filter-seqs \
        --i-data $data_file \
        --i-table $table_file \
        --o-filtered-data ${data_file%.qza}-filtered.qza
done

```




## 4. RESULTS FROM DIFFERENT METHODS

**4.1. Results from trimmming and clustering (asv's and otus approach)**

```{r, warning=FALSE, message=FALSE, eval=FALSE}

#loading files and package to summarize
library(tidyverse)
library(qiime2R)
library(plyr)
library(RColorBrewer)


#loadding tables to summarice
trimmed_table<- data.frame(t(read_qza("../COI/TRIMMED/demux_trimmed/demux_lib/no_cluster/table_trimmed_paired.qza")$data), check.names = F )
trimmed_clustered_table<-data.frame(t(read_qza("../COI/TRIMMED/demux_trimmed/demux_lib/cluster/table_clustered.qza")$data), check.names = F) 

trimmed_discard_table<- data.frame(t(read_qza("../COI/TRIMMED/demux_trimmed_discard/demux_lib/tables/table_trimmed_paired_discard.qza")$data) , check.names = F )
trimmed_discard_clustered_table<-data.frame(t(read_qza("../COI/TRIMMED/demux_trimmed_discard/demux_lib/clustered/table_clustered_disc.qza")$data) , check.names = F )

trimmed_comple_table<-  data.frame(t(read_qza("../COI/TRIMMED/demux_trimmed_complementarios/no_clustered/table_complementarios.qza")$data)  , check.names = F )
trimmed_comple_clustered_table<-data.frame(t(read_qza("../COI/TRIMMED/demux_trimmed_complementarios/clustered/table_clustered_comple.qza")$data)  , check.names = F )

trimmed_comple_discard_table<-data.frame(t(read_qza("../COI/TRIMMED/demux_trimmed_complementarios_discard/no_clustered/table_complementarios_discard.qza")$data), check.names = F )
trimmed_comple_discard_clustered_table<-data.frame(t(read_qza("../COI/TRIMMED/demux_trimmed_complementarios_discard/clustered/table_clustered_comple_disc.qza")$data ), check.names = F )

otus_table<- data.frame(t(read_qza("../COI/TRIMMED/demux_trimmed/demux_lib/otus_strategy/table_otus_filter_singletons.qza")$data) , check.names = F )



#loading files to summarize
#these files were obtained in qiime2 (e.g for trimmed table "qiime tools export --input-path table_trimmed.qzv --output-path trimmed")

trimmed<- read_csv("../COI/TRIMMED/demux_trimmed/demux_lib/no_cluster/trimmed/feature-frequency-detail.csv")%>% dplyr::select(-X1, frequency=`0`)
trimmed_clustered<- read_csv("../COI/TRIMMED/demux_trimmed/demux_lib/cluster/trimmed_clustered/feature-frequency-detail.csv") %>% dplyr::select(-X1, frequency=`0`)

trimmed_discard<- read_csv("../COI/TRIMMED/demux_trimmed_discard/demux_lib/no_clustered/trimmed_discard/feature-frequency-detail.csv")%>% dplyr::select(-X1, frequency=`0`)
trimmed_discard_clustered<-read_csv("../COI/TRIMMED/demux_trimmed_discard/demux_lib/clustered/trimmed_discard_clustered/feature-frequency-detail.csv")%>% dplyr::select(-X1, frequency=`0`)

trimmed_comple<- read_csv("../COI/TRIMMED/demux_trimmed_complementarios/no_clustered/trimmed_complementarios/feature-frequency-detail.csv")%>% dplyr::select(-X1, frequency=`0`)
trimmed_comple_clustered<- read_csv("../COI/TRIMMED/demux_trimmed_complementarios/clustered/trimmed_clustered_complementarios/feature-frequency-detail.csv")%>% dplyr::select(-X1, frequency=`0`)

trimmed_comple_discard<- read_csv("../COI/TRIMMED/demux_trimmed_complementarios_discard/no_clustered/trimmed_comple_discard/feature-frequency-detail.csv")%>% dplyr::select(-X1, frequency=`0`)
trimmed_comple_discard_clustered<-read_csv("../COI/TRIMMED/demux_trimmed_complementarios_discard/clustered/trimmed_comple_discad/feature-frequency-detail.csv")%>% dplyr::select(-X1, frequency=`0`)

otus<-read_csv("../COI/TRIMMED/demux_trimmed/demux_lib/otus_strategy/otus/feature-frequency-detail.csv")%>% dplyr::select(-X1, frequency=`0`)


```

```{r, warning=FALSE, message=FALSE, eval=FALSE}

#Let's summarice the uniques features (e.g, features that are in just one sample)

tables_all<- list(trimmed_table, trimmed_clustered_table, trimmed_discard_table, trimmed_discard_clustered_table,trimmed_comple_table, trimmed_comple_clustered_table, trimmed_comple_discard_table, trimmed_comple_discard_clustered_table,otus_table)


nonzero <- function(x) sum(x != 0)
count_nonzero<-function(y) {numcolwise(nonzero)(y) %>% t() %>% as.data.frame() %>% filter(V1 == 1)%>% nrow()}

tables_uniques<- lapply(tables_all, count_nonzero)

names_rows_uniques<- c("trimmed", "trimmed_clustered", "trimmed_discard", "trimmed_discard_clustered", "trimmed_comple", "trimmed_comple_clustered", "trimmed_comple_discard", "trimmed_comple_discard_clustered", "otus")


table_summaricing_uniques  <-  as.data.frame(matrix(unlist(tables_uniques), nrow=9)) %>%dplyr::select("Features just one sample" = "V1")
rownames(table_summaricing_uniques)<- names_rows_uniques

  
```


```{r, message=FALSE, warning=FALSE, eval=FALSE}
#let's summarice the total of features and counts in each method
library(kableExtra)
tables_sum<- list(trimmed, trimmed_clustered, trimmed_discard, trimmed_discard_clustered,
                  trimmed_comple, trimmed_comple_clustered, trimmed_comple_discard, 
                  trimmed_comple_discard_clustered,otus)


names_rows<- c("trimmed", "trimmed_clustered", "trimmed_discard", "trimmed_discard_clustered", "trimmed_comple", "trimmed_comple_clustered", "trimmed_comple_discard", "trimmed_comple_discard_clustered", "otus")

summarice_freq <- function(x){
  list(colSums(x),nrow(x))
}
summaricing_freq<-sapply(tables_sum, summarice_freq )
dd  <-  as.data.frame(matrix(unlist(summaricing_freq), nrow=2)) 
table_summaricing<-dd %>% t() %>% as.data.frame() %>% dplyr::rename( "Total frequency" = "V1", "Number of features"="V2")
rownames(table_summaricing)<- names_rows

table_sum<- cbind(table_summaricing, table_summaricing_uniques)


kbl(table_sum) %>% kable_paper(bootstrap_options = "striped", full_width = F, position = "center")

```



**4.2. PCA plots**


```{r, warning=FALSE, message=FALSE, message=FALSE, eval=FALSE}

# load the required R packages and files 
require(compositions) 
require(zCompositions)
library(CoDaSeq)

metadata<-read_tsv("../COI/MAPPINGS/merge_mapping.txt")

metadata$Ecosystems<- factor(metadata$Ecosystem,
                         levels = c( "Bosque", "Cultivo", "PA"),
                         labels = c("Forest", "Cultivation" ,"PA"))
metadata$sampleid<- as.character(metadata$sampleid)

#setting functions 
zero_transform<- function(x){data.frame(t(cmultRepl(t(x), method="CZM", output="p-counts")), check.names = F)}
d.clr.abund.codaseq<-function(x){data.frame(codaSeq.clr(x,samples.by.row = F), check.names = F)}
pcx.abunda <- function(x){prcomp((x))}

#applying functions
zero_tables<- lapply(tables_all, zero_transform)
d.clr.abund<- lapply(zero_tables, d.clr.abund.codaseq)
pcx.abund<- lapply(d.clr.abund, pcx.abunda)


```

```{r, warning=FALSE, message=FALSE,  fig.show="hold", out.width="50%", message=FALSE, eval=FALSE}

#create the base plot with only the arrows
  PC1.f<- function(x,y){paste("PC1", round(sum(x$sdev[1] ^ 2) / mvar(y) * 100, 1), "%")}
  PC2.f <- function(x,y){paste("PC2", round(sum(x$sdev[2] ^ 2) / mvar(y) * 100, 1), "%")}
  
  
  PC1_all<- mapply(PC1.f, pcx.abund, d.clr.abund)
  PC2_all<- mapply(PC2.f, pcx.abund, d.clr.abund)
  
  pca_tables<- function(tab){ggplot() +
   # theme_q2r() +
    #xlab(PC1) +
    #ylab(PC2) +
   geom_segment(data=data.frame(tab$rotation) %>%   #arrows
                   rownames_to_column(var = "FeatureID")%>%  
                   mutate(a=sqrt(PC1^2+PC2^2)) %>% # calculate the distance from the origin
                   top_n(15, a) %>% #keep 10 furthest away points
                   mutate(PC1=PC1*800, PC2=PC2*800),
                   aes(x=0, xend=PC1, y=0, yend=PC2),
                   arrow = arrow(length = unit(0.3,"cm")))+
    geom_point(data=data.frame(tab$x) %>% #individuals
                   rownames_to_column(var = "sampleid")%>%
                   left_join(metadata, by = "sampleid"),
                   aes(x=PC1, y=PC2, fill=Ecosystems),shape=21, size=4) +
                   geom_vline(xintercept = 0, linetype = 2) +   #lines-cross
                   geom_hline(yintercept = 0, linetype = 2) +
                  #scale_fill_manual(values = c("#212F3D","#839192"))#color of points 
     theme(axis.text = element_text(colour = "black", size = 12),
                   axis.title = element_text(colour = "black", size = 12),
                   legend.text = element_text(size = 10),
                   legend.title = element_text(size = 12), 
                   legend.position = "right", 
                   legend.box = "vertical") }
  
  
  
  
  figures_pca<- lapply(pcx.abund, pca_tables)
  
  pca_trimmed<- figures_pca[[1]]+ xlab(PC1_all[[1]]) +ylab(PC2_all[[1]]) + ggtitle("trimmed")
  pca_trimmed_clustered<- figures_pca[[2]]+ xlab(PC1_all[[2]]) +ylab(PC2_all[[2]]) + ggtitle("trimmed clustered")
pca_trimmed_discard<- figures_pca[[3]]+ xlab(PC1_all[[3]]) +ylab(PC2_all[[3]]) + ggtitle("trimmed discard")
pca_trimmed_discard_clustered<- figures_pca[[4]]+ xlab(PC1_all[[4]]) +ylab(PC2_all[[4]]) + ggtitle("trimmed discard clustered")
pca_trimmed_comple<- figures_pca[[5]]+ xlab(PC1_all[[5]]) +ylab(PC2_all[[5]]) + ggtitle("trimmed complementaries")
pca_trimmed_comple_clustered<- figures_pca[[6]]+ xlab(PC1_all[[6]]) +ylab(PC2_all[[6]]) + ggtitle("trimmed complementaries clustered")
pca_trimmed_comple_discard<- figures_pca[[7]]+ xlab(PC1_all[[7]]) +ylab(PC2_all[[7]]) + ggtitle("trimmed complementaries discard")
pca_trimmed_ccomple_discard_clustered<- figures_pca[[8]]+ xlab(PC1_all[[8]]) +ylab(PC2_all[[8]]) + ggtitle("trimmed complementaries discard clustered")
pca_otus<- figures_pca[[9]]+ xlab(PC1_all[[9]]) +ylab(PC2_all[[9]]) + ggtitle("otus approach")

pca_trimmed;pca_trimmed_clustered; pca_trimmed_discard; pca_trimmed_discard_clustered; pca_trimmed_comple; pca_trimmed_comple_clustered; pca_trimmed_comple_discard; pca_trimmed_ccomple_discard_clustered;pca_otus

```

**4.3. Colapsing to Genus**

```{r, warning=FALSE, message=FALSE, fig.show="hold", out.width="50%", message=FALSE, eval=FALSE}

#I chose trimmed clustered, no-clustered and otus approach

table_genus_clustered<- data.frame(t(read_qza("../COI/TRIMMED/demux_trimmed/demux_lib/cluster/table_level7.qza")$data) , check.names = F )
table_genus_no_clustered<- data.frame(t(read_qza("../COI/TRIMMED/demux_trimmed/demux_lib/no_cluster//table_level7_noclustered.qza")$data) , check.names = F )
table_genus_otus<- data.frame(t(read_qza("../COI/clustered_data/table_otus_level7.qza")$data) , check.names = F )


metadata<-read_tsv("../COI/MAPPINGS/merge_mapping.txt")

metadata$Ecosystems<- factor(metadata$Ecosystem,
                             levels = c( "Bosque", "Cultivo", "PA"),
                             labels = c("Forest", "Cultivation" ,"PA"))
metadata$sampleid<- as.character(metadata$sampleid)



#We use the same functions we used before (for PCA's)
tables_all2<- list(table_genus_clustered, table_genus_no_clustered, table_genus_otus)

zero_tables<- lapply(tables_all2, zero_transform)
d.clr.abund<- lapply(zero_tables, d.clr.abund.codaseq)
pcx.abund<- lapply(d.clr.abund, pcx.abunda)


PC1_all<- mapply(PC1.f, pcx.abund, d.clr.abund)
PC2_all<- mapply(PC2.f, pcx.abund, d.clr.abund)

pca_tables<- function(tab){ggplot() +
    # theme_q2r() +
    #xlab(PC1) +
    #ylab(PC2) +
    geom_segment(data=data.frame(tab$rotation) %>%   #arrows
                   rownames_to_column(var = "FeatureID")%>%  
                   mutate(a=sqrt(PC1^2+PC2^2)) %>% # calculate the distance from the origin
                   top_n(15, a) %>% #keep 10 furthest away points
                   mutate(PC1=PC1*10, PC2=PC2*10),
                 aes(x=0, xend=PC1, y=0, yend=PC2),
                 arrow = arrow(length = unit(0.3,"cm")))+
    geom_point(data=data.frame(tab$x) %>% #individuals
                 rownames_to_column(var = "sampleid")%>%
                 left_join(metadata, by = "sampleid"),
               aes(x=PC1, y=PC2, color=Altitud, shape= Ecosystems), size=4) +
    scale_color_viridis_c()+
    geom_vline(xintercept = 0, linetype = 2) +   #lines-cross
    geom_hline(yintercept = 0, linetype = 2) +
    #scale_fill_manual(values = c("#212F3D","#839192"))#color of points 
    theme(axis.text = element_text(colour = "black", size = 12),
          axis.title = element_text(colour = "black", size = 12),
          legend.text = element_text(size = 10),
          legend.title = element_text(size = 12), 
          legend.position = "right", 
          legend.box = "vertical") }




figures_pca<- lapply(pcx.abund, pca_tables)

pca_genus_clustered<- figures_pca[[1]]+ xlab(PC1_all[[1]]) +ylab(PC2_all[[1]]) + ggtitle("clustered")
pca_genus_noclustered<- figures_pca[[2]]+ xlab(PC1_all[[2]]) +ylab(PC2_all[[2]]) + ggtitle("no clustered")
pca_genus_otus<- figures_pca[[3]]+ xlab(PC1_all[[3]]) +ylab(PC2_all[[3]]) + ggtitle("otus")

pca_genus_clustered; pca_genus_noclustered;pca_genus_otus

```

**4.4.Choosing a taxa to cluster**

```{r, eval=FALSE}

table_phylum_clustered<- data.frame(t(read_qza("../COI/TRIMMED/demux_trimmed/demux_lib/cluster/table_level2_clustered.qza")$data) , check.names = F )
table_phylum_no_clustered<- data.frame(t(read_qza("../COI/TRIMMED/demux_trimmed/demux_lib/no_cluster//table_level2_noclustered.qza")$data) , check.names = F )
table_phylum_otus<- data.frame(t(read_qza("../COI/clustered_data/table_otus_level2.qza")$data) , check.names = F )

table_phylum_clustered_gather<- table_phylum_clustered %>%   gather(key = "taxa", value = "clustered")%>% group_by(taxa) %>% summarise_if(is.numeric, sum) 

table_phylum_no_clustered_gather<- table_phylum_no_clustered %>%   gather(key = "taxa", value = "no_clustered")%>% group_by(taxa) %>% summarise_if(is.numeric, sum) 

table_phylum_otus_gather<- table_phylum_otus %>%   gather(key = "taxa", value = "otus")%>% group_by(taxa) %>% summarise_if(is.numeric, sum) 

tables_phylum<- table_phylum_clustered_gather %>% full_join(table_phylum_no_clustered_gather, by = "taxa") %>% full_join(table_phylum_otus_gather, by = "taxa")%>% replace(is.na(.), 0)


  
  tables_phylum %>% gather(-taxa, key = "ids", value = "counts") %>% ggplot(aes(ids, counts, fill=taxa)) +
  geom_bar( stat = "identity", position = "fill") +
  scale_fill_manual(name = "Taxa",
                    values = c(brewer.pal(12, "Paired")[1:12],
                               brewer.pal(8, "Set2")[1:8], brewer.pal(12, "Paired")[1:12]))

```


```
As we see the major percentage are assigned to Arthropoa phylum

```
**4.5. Arthropoda tables**

```{r, warning=FALSE, message=FALSE,fig.show="hold", out.width="50%", message=FALSE, eval=FALSE}
#Tables filter to just keep Arthropoda phylum using qiime2 

table_arthropoda_clustered<-data.frame(t(read_qza("../COI/TRIMMED/demux_trimmed/demux_lib/cluster/table_arthropoda.qza")$data) , check.names = F )
table_arthropoda_no_clustered<- data.frame(t(read_qza("../COI/TRIMMED/demux_trimmed/demux_lib/no_cluster/table_arthropoda.qza")$data) , check.names = F )
table_arthropoda_otus<- data.frame(t(read_qza("../COI/clustered_data/table_arthropoda.qza")$data) , check.names = F )

table_arthropoda_clustered_genus<-data.frame(t(read_qza("../COI/TRIMMED/demux_trimmed/demux_lib/cluster/table_level7_clustered_arto.qza")$data) , check.names = F )
table_arthropoda_no_clustered_genus<- data.frame(t(read_qza("../COI/TRIMMED/demux_trimmed/demux_lib/no_cluster/table_level7_clustered_arto.qza")$data) , check.names = F )
table_arthropoda_otus_genus<- data.frame(t(read_qza("../COI/clustered_data/table_otus_level7_arto.qza")$data) , check.names = F )

tables_all3<- list(table_arthropoda_clustered, table_arthropoda_no_clustered, 
                   table_arthropoda_otus,table_arthropoda_clustered_genus,
                   table_arthropoda_no_clustered_genus, table_arthropoda_otus_genus)


zero_tables<- lapply(tables_all3, zero_transform)
d.clr.abund<- lapply(zero_tables, d.clr.abund.codaseq)
pcx.abund<- lapply(d.clr.abund, pcx.abunda)

PC1_all<- mapply(PC1.f, pcx.abund, d.clr.abund)
PC2_all<- mapply(PC2.f, pcx.abund, d.clr.abund)

pca_tables<- function(tab){ggplot() +
    # theme_q2r() +
    #xlab(PC1) +
    #ylab(PC2) +
    geom_segment(data=data.frame(tab$rotation) %>%   #arrows
                   rownames_to_column(var = "FeatureID")%>%  
                   mutate(a=sqrt(PC1^2+PC2^2)) %>% # calculate the distance from the origin
                   top_n(15, a) %>% #keep 10 furthest away points
                   mutate(PC1=PC1*10, PC2=PC2*10),
                 aes(x=0, xend=PC1, y=0, yend=PC2),
                 arrow = arrow(length = unit(0.3,"cm")))+
    geom_point(data=data.frame(tab$x) %>% #individuals
                 rownames_to_column(var = "sampleid")%>%
                 left_join(metadata, by = "sampleid"),
               aes(x=PC1, y=PC2, color=Altitud, shape= Ecosystems), size=4) +
    scale_color_viridis_c(option = "plasma")+
    geom_vline(xintercept = 0, linetype = 2) +   #lines-cross
    geom_hline(yintercept = 0, linetype = 2) +
    #scale_fill_manual(values = c("#212F3D","#839192"))#color of points 
    theme(axis.text = element_text(colour = "black", size = 12),
          axis.title = element_text(colour = "black", size = 12),
          legend.text = element_text(size = 10),
          legend.title = element_text(size = 12), 
          legend.position = "right", 
          legend.box = "vertical") }

  
figures_pca<- lapply(pcx.abund, pca_tables)

pca_clustered_arto<- figures_pca[[1]]+ xlab(PC1_all[[1]]) +ylab(PC2_all[[1]]) + ggtitle("clustered")
pca_noclustered_arto<- figures_pca[[2]]+ xlab(PC1_all[[2]]) +ylab(PC2_all[[2]]) + ggtitle("no clustered")
pca_otus_arto<- figures_pca[[3]]+ xlab(PC1_all[[3]]) +ylab(PC2_all[[3]]) + ggtitle("otus")

pca_genus_clustered_arto<- figures_pca[[4]]+ xlab(PC1_all[[4]]) +ylab(PC2_all[[4]]) + ggtitle("clustered_genus")
pca_genus_noclustered_arto<- figures_pca[[5]]+ xlab(PC1_all[[5]]) +ylab(PC2_all[[5]]) + ggtitle("no clustered_genus")
pca_genus_otus_arto<- figures_pca[[6]]+ xlab(PC1_all[[6]]) +ylab(PC2_all[[6]]) + ggtitle("otus_genus")

pca_clustered_arto; pca_noclustered_arto; pca_otus_arto; pca_genus_clustered_arto; pca_genus_noclustered_arto;pca_genus_otus_arto


```

**4.5.Collapsing to Genus but with allbolDB taxonomy**

```{r, warning=FALSE, message=FALSE,  fig.show="hold", out.width="50%", message=FALSE, eval=FALSE}

#allboldDB taxonomy is set to focus more in Arthropoda animalia taxa

table_allbold_clustered_genus<-data.frame(t(read_qza("../COI/clustered_data/table_clustered_level7_allbold.qza")$data) , check.names = F )
table_allbold_no_clustered_genus<- data.frame(t(read_qza("../COI/cluster_data_no/table_noclustered_level7_allbold.qza")$data) , check.names = F )
table_allbold_otus_genus<- data.frame(t(read_qza("../COI/clustered_data/table_otus_level7_allbold.qza")$data) , check.names = F )

tables_all4<- list(table_allbold_clustered_genus, table_allbold_no_clustered_genus, table_allbold_otus_genus)


zero_tables<- lapply(tables_all4, zero_transform)
d.clr.abund<- lapply(zero_tables, d.clr.abund.codaseq)
pcx.abund<- lapply(d.clr.abund, pcx.abunda)

PC1_all<- mapply(PC1.f, pcx.abund, d.clr.abund)
PC2_all<- mapply(PC2.f, pcx.abund, d.clr.abund)

pca_tables<- function(tab){ggplot() +
    # theme_q2r() +
    #xlab(PC1) +
    #ylab(PC2) +
    geom_segment(data=data.frame(tab$rotation) %>%   #arrows
                   rownames_to_column(var = "FeatureID")%>%  
                   mutate(a=sqrt(PC1^2+PC2^2)) %>% # calculate the distance from the origin
                   top_n(15, a) %>% #keep 10 furthest away points
                   mutate(PC1=PC1*10, PC2=PC2*10),
                 aes(x=0, xend=PC1, y=0, yend=PC2),
                 arrow = arrow(length = unit(0.3,"cm")))+
    geom_point(data=data.frame(tab$x) %>% #individuals
                 rownames_to_column(var = "sampleid")%>%
                 left_join(metadata, by = "sampleid"),
               aes(x=PC1, y=PC2, color=Altitud, shape= Ecosystems), size=4) +
    scale_color_viridis_c(option = "inferno")+
    geom_vline(xintercept = 0, linetype = 2) +   #lines-cross
    geom_hline(yintercept = 0, linetype = 2) +
    #scale_fill_manual(values = c("#212F3D","#839192"))#color of points 
    theme(axis.text = element_text(colour = "black", size = 12),
          axis.title = element_text(colour = "black", size = 12),
          legend.text = element_text(size = 10),
          legend.title = element_text(size = 12), 
          legend.position = "right", 
          legend.box = "vertical") }

  
figures_pca<- lapply(pcx.abund, pca_tables)

pca_clustered_allbold<- figures_pca[[1]]+ xlab(PC1_all[[1]]) +ylab(PC2_all[[1]]) + ggtitle("clustered")
pca_noclustered_allbold<- figures_pca[[2]]+ xlab(PC1_all[[2]]) +ylab(PC2_all[[2]]) + ggtitle("no clustered")
pca_otus_allbold<- figures_pca[[3]]+ xlab(PC1_all[[3]]) +ylab(PC2_all[[3]]) + ggtitle("otus")

pca_clustered_allbold; pca_noclustered_allbold; pca_otus_allbold


```


